---
title: "Fake Pramipexole Trial Simulation"
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(lme4)
library(lmerTest)
library(emmeans)
library(broom.mixed)
library(knitr)

set.seed(20260311)
```

## Goal

This Quarto document generates a **fully fictitious dataset** inspired by the design and the main orders of magnitude reported in `article.pdf`.

Important points:

- These are **not the real trial data**
- The simulated sample size follows your request: **75 patients in the pramipexole arm and 70 in the placebo arm**
- The real article used a **1:1 randomisation** and had **75 vs 75 eligible randomised participants**
- The analysis below is only an **illustration** of a plausible mixed-model workflow close to the published primary analysis

## Simulation assumptions

The fake data are loosely based on the paper:

- mean baseline QIDS-SR16 around **16.3**
- age around **44.9 years**
- about **56% women**
- 9 sites with a distribution roughly inspired by the published table
- a stronger reduction in QIDS-SR16 in the pramipexole arm by week 12
- gradual increase in missing data over follow-up

## 1. Simulate participant-level baseline data

```{r}
sites <- c(
  "Oxford", "Newcastle", "Liverpool", "King's College London",
  "Bristol", "UCL", "Nottingham", "Midlands Partnership", "Pennine Care"
)

site_probs <- c(74, 26, 12, 12, 10, 6, 6, 3, 1)
site_probs <- site_probs / sum(site_probs)

n_prami <- 75
n_placebo <- 70
n_total <- n_prami + n_placebo

draw_qids_baseline <- function(severity_cat) {
  if (severity_cat == "11-15") {
    x <- rnorm(1, mean = 13.5, sd = 1.2)
    return(min(max(x, 11), 15))
  }
  if (severity_cat == "16-20") {
    x <- rnorm(1, mean = 17.6, sd = 1.2)
    return(min(max(x, 16), 20))
  }
  x <- rnorm(1, mean = 22.0, sd = 1.3)
  min(max(x, 20.1), 25)
}

participants <- tibble(
  id = 1:n_total,
  site = sample(sites, size = n_total, replace = TRUE, prob = site_probs),
  age = pmin(pmax(rnorm(n_total, mean = 44.9, sd = 14), 18), 74),
  sex = sample(c("Female", "Male"), size = n_total, replace = TRUE, prob = c(0.56, 0.44)),
  severity_cat = sample(
    c("11-15", "16-20", ">20"),
    size = n_total,
    replace = TRUE,
    prob = c(0.45, 0.43, 0.12)
  )
) |>
  mutate(
    age_group = if_else(age <= 50, "18-50", ">50"),
    baseline_qids = vapply(severity_cat, draw_qids_baseline, numeric(1))
  ) |>
  slice_sample(prop = 1) |>
  mutate(entry_order = row_number())

kable(head(participants, 10), digits = 2, caption = "First 10 fictitious participants")
```

## 2. Assign treatment with a minimisation algorithm

This section mimics the published randomisation logic:

- first 10 participants: random allocation without minimisation
- then a minimisation algorithm
- 80% probability in favour of the allocation that minimises imbalance

To match your request, the fake dataset is capped at **75 pramipexole** and **70 placebo**.

```{r}
imbalance_score <- function(df) {
  score_vars <- c("site", "age_group", "sex", "severity_cat")
  total_score <- 0

  for (v in score_vars) {
    tab <- table(df[[v]], df$treatment)
    for (arm in c("Placebo", "Pramipexole")) {
      if (!arm %in% colnames(tab)) {
        tab <- cbind(tab, setNames(rep(0, nrow(tab)), arm))
      }
    }
    tab <- tab[, c("Placebo", "Pramipexole"), drop = FALSE]
    total_score <- total_score + sum(abs(tab[, "Placebo"] - tab[, "Pramipexole"]))
  }

  total_score
}

assignments <- participants
assignments$treatment <- NA_character_

for (i in seq_len(n_total)) {
  current <- assignments[seq_len(i - 1), , drop = FALSE]
  remaining_prami <- n_prami - sum(current$treatment == "Pramipexole", na.rm = TRUE)
  remaining_placebo <- n_placebo - sum(current$treatment == "Placebo", na.rm = TRUE)

  if (remaining_prami == 0) {
    assignments$treatment[i] <- "Placebo"
    next
  }
  if (remaining_placebo == 0) {
    assignments$treatment[i] <- "Pramipexole"
    next
  }

  if (i <= 10) {
    assignments$treatment[i] <- sample(c("Placebo", "Pramipexole"), size = 1)
    next
  }

  candidate_placebo <- bind_rows(current, mutate(assignments[i, ], treatment = "Placebo"))
  candidate_prami   <- bind_rows(current, mutate(assignments[i, ], treatment = "Pramipexole"))

  score_placebo <- imbalance_score(candidate_placebo)
  score_prami   <- imbalance_score(candidate_prami)

  if (score_placebo < score_prami) {
    favoured <- "Placebo"
  } else if (score_prami < score_placebo) {
    favoured <- "Pramipexole"
  } else {
    favoured <- sample(c("Placebo", "Pramipexole"), size = 1)
  }

  if (favoured == "Pramipexole") {
    assignments$treatment[i] <- sample(c("Pramipexole", "Placebo"), size = 1, prob = c(0.8, 0.2))
  } else {
    assignments$treatment[i] <- sample(c("Placebo", "Pramipexole"), size = 1, prob = c(0.8, 0.2))
  }
}

participants <- assignments |>
  mutate(
    treatment = factor(treatment, levels = c("Placebo", "Pramipexole")),
    age_group = factor(age_group, levels = c("18-50", ">50")),
    sex = factor(sex, levels = c("Female", "Male")),
    severity_cat = factor(severity_cat, levels = c("11-15", "16-20", ">20")),
    site = factor(site)
  )

participants |>
  count(treatment) |>
  kable(caption = "Final fictitious allocation counts")
```

```{r}
baseline_summary <- participants |>
  group_by(treatment) |>
  summarise(
    n = n(),
    age_mean = mean(age),
    female_pct = mean(sex == "Female") * 100,
    baseline_qids_mean = mean(baseline_qids),
    severe_or_very_severe_pct = mean(severity_cat %in% c("16-20", ">20")) * 100,
    .groups = "drop"
  )

kable(baseline_summary, digits = 2, caption = "Baseline summary by arm")
```

## 3. Simulate repeated QIDS-SR16 measurements

```{r}
weeks <- c(2, 4, 6, 8, 10, 12, 16, 20, 24, 28, 32, 36, 40, 44, 48)

placebo_change <- c(-0.4, -0.9, -1.3, -1.7, -2.1, -2.5,
                    -2.8, -3.1, -3.3, -3.5, -3.7, -3.8, -3.9, -4.0, -4.1)

prami_extra <- c(-0.5, -1.1, -1.8, -2.4, -3.1, -3.9,
                 -3.6, -3.4, -3.1, -2.8, -2.6, -2.4, -2.2, -2.1, -2.0)

site_re <- rnorm(length(sites), mean = 0, sd = 0.35)
names(site_re) <- sites

participants <- participants |>
  mutate(
    site_re = site_re[as.character(site)],
    subject_intercept = rnorm(n_total, mean = 0, sd = 1.2),
    subject_slope = rnorm(n_total, mean = 0, sd = 0.8)
  )

longitudinal <- tidyr::expand_grid(id = participants$id, week = weeks) |>
  left_join(participants, by = "id") |>
  mutate(
    placebo_component = placebo_change[match(week, weeks)],
    treatment_component = if_else(treatment == "Pramipexole", prami_extra[match(week, weeks)], 0),
    subject_time_component = subject_slope * sqrt(week / 12),
    age_component = if_else(age_group == ">50", 0.3, 0),
    sex_component = if_else(sex == "Male", 0.1, 0),
    qids = baseline_qids +
      placebo_component +
      treatment_component +
      age_component +
      sex_component +
      site_re +
      subject_intercept * 0.25 +
      subject_time_component +
      rnorm(n(), mean = 0, sd = 1.5),
    qids = pmin(pmax(qids, 0), 27)
  )
```

## 4. Add plausible missingness

```{r}
sample_dropout_week <- function(trt) {
  stage <- if (trt == "Pramipexole") {
    sample(c("pre12", "post12", "complete"), size = 1, prob = c(0.10, 0.40, 0.50))
  } else {
    sample(c("pre12", "post12", "complete"), size = 1, prob = c(0.10, 0.38, 0.52))
  }

  if (stage == "pre12") {
    return(sample(c(2, 4, 6, 8, 10), size = 1))
  }
  if (stage == "post12") {
    return(sample(c(12, 16, 20, 24, 28, 32, 36, 40, 44), size = 1))
  }
  Inf
}

dropout_df <- participants |>
  transmute(
    id,
    dropout_week = vapply(as.character(treatment), sample_dropout_week, numeric(1))
  )

longitudinal <- longitudinal |>
  left_join(dropout_df, by = "id") |>
  mutate(
    observed = week <= dropout_week,
    observed = if_else(observed, runif(n()) > 0.03, FALSE)
  )

analysis_df <- longitudinal |>
  filter(observed) |>
  mutate(week_f = factor(week))

availability <- analysis_df |>
  filter(week %in% c(12, 48)) |>
  count(treatment, week) |>
  tidyr::pivot_wider(names_from = week, values_from = n, names_prefix = "week_")

kable(availability, caption = "Observed data counts at week 12 and week 48")
```

## 5. Save the fake data frames

```{r}
participant_export <- participants |>
  select(id, treatment, site, age, age_group, sex, severity_cat, baseline_qids)

long_export <- analysis_df |>
  select(id, treatment, site, age_group, sex, severity_cat, baseline_qids, week, qids)

write.csv(participant_export, "fake_trial_participants.csv", row.names = FALSE)
write.csv(long_export, "fake_trial_longitudinal.csv", row.names = FALSE)
```

The following files are written in the working directory:

- `fake_trial_participants.csv`
- `fake_trial_longitudinal.csv`

## 6. Descriptive results from the fictitious data

```{r}
observed_summary <- analysis_df |>
  filter(week %in% c(12, 48)) |>
  mutate(change_from_baseline = qids - baseline_qids) |>
  group_by(treatment, week) |>
  summarise(
    n = n(),
    mean_qids = mean(qids),
    mean_change = mean(change_from_baseline),
    sd_change = sd(change_from_baseline),
    .groups = "drop"
  )

kable(observed_summary, digits = 2, caption = "Observed fake outcomes at week 12 and week 48")
```

## 7. Fit an approximate mixed model

This model is designed to stay close to the published analysis:

```{r}
model_formula <- qids ~ treatment * week_f + age_group + sex + severity_cat + baseline_qids + (1 | site) + (1 | id)
model_formula
```

```{r}
fit <- lmerTest::lmer(
  formula = model_formula,
  data = analysis_df,
  REML = FALSE
)
```

## 8. How an adjusted mean difference is obtained

For a continuous outcome, the paper reports an **adjusted mean difference** rather
than a raw difference in observed means.

In a linear mixed model, this means:

- the model first estimates the average QIDS-SR16 trajectory after adjustment for
  treatment, time, baseline QIDS-SR16, age group, sex, severity category, site,
  and participant-level clustering
- then the model-derived mean for each treatment arm is obtained at a given timepoint
- the **adjusted mean difference** is the contrast between these two model-based means

In this coding, with week treated as a factor, the adjusted difference at week 12 is:

```text
Adjusted MD at week 12
= β[treatmentPramipexole] + β[treatmentPramipexole:week_f12]
```

because:

```text
Predicted mean at week 12 in placebo
= β0 + β[week_f12] + covariate terms

Predicted mean at week 12 in pramipexole
= β0 + β[treatmentPramipexole] + β[week_f12]
  + β[treatmentPramipexole:week_f12] + covariate terms
```

and the common covariate terms cancel out in the contrast.

### Raw week-12 difference vs adjusted week-12 difference

```{r}
raw_week12 <- analysis_df |>
  filter(week == 12) |>
  group_by(treatment) |>
  summarise(
    n = n(),
    mean_qids = mean(qids),
    .groups = "drop"
  )

raw_md_week12 <- with(raw_week12, mean_qids[treatment == "Pramipexole"] - mean_qids[treatment == "Placebo"])

raw_week12 |>
  kable(digits = 3, caption = "Observed means at week 12")

cat("\nObserved raw mean difference at week 12 (Pramipexole - Placebo):",
    round(raw_md_week12, 3), "\n")
```

```{r}
emm_all <- emmeans::emmeans(fit, ~ treatment | week_f)

emm_week12 <- summary(emm_all) |>
  as.data.frame() |>
  filter(week_f == "12")

contrast_week12 <- pairs(emm_all, reverse = TRUE) |>
  summary(infer = c(TRUE, TRUE)) |>
  as.data.frame() |>
  filter(week_f == "12")

kable(emm_week12, digits = 3, caption = "Model-based adjusted means at week 12")
kable(contrast_week12, digits = 3, caption = "Adjusted mean difference at week 12")
```

### Direct check from the fixed-effect coefficients

```{r}
coefs <- lme4::fixef(fit)

coef_check <- tibble(
  term = c("treatmentPramipexole", "treatmentPramipexole:week_f12"),
  estimate = c(
    unname(coefs["treatmentPramipexole"]),
    unname(coefs["treatmentPramipexole:week_f12"])
  )
) |>
  bind_rows(
    tibble(
      term = "sum = adjusted mean difference at week 12",
      estimate = sum(coefs["treatmentPramipexole"], coefs["treatmentPramipexole:week_f12"])
    )
  )

kable(coef_check, digits = 3, caption = "Coefficient decomposition of the adjusted week-12 contrast")
```

### Random-effects standard deviations

```{r}
ran_pars <- broom.mixed::tidy(fit, effects = "ran_pars", scales = "sdcor")
kable(ran_pars, digits = 3, caption = "Random-effects parameters")
```

### Adjusted treatment contrasts at week 12 and week 48

```{r}
contrasts_by_week <- pairs(emm_all, reverse = TRUE) |>
  broom::tidy(conf.int = TRUE) |>
  mutate(week = as.integer(as.character(week_f))) |>
  filter(week %in% c(12, 48)) |>
  select(week, contrast, estimate, std.error, statistic, p.value, conf.low, conf.high)

kable(contrasts_by_week, digits = 3, caption = "Adjusted Pramipexole - Placebo contrasts")
```

### Selected fixed effects

```{r}
fixed_effects <- broom.mixed::tidy(fit, effects = "fixed", conf.int = TRUE) |>
  filter(grepl("treatmentPramipexole|week_f12|treatmentPramipexole:week_f12|week_f48|treatmentPramipexole:week_f48|baseline_qids|age_group|sex|severity_cat", term))

kable(fixed_effects, digits = 3, caption = "Selected fixed effects from the mixed model")
```

## 9. Illustration plot

```{r}
plot_df <- analysis_df |>
  group_by(treatment, week) |>
  summarise(
    mean_qids = mean(qids),
    se = sd(qids) / sqrt(n()),
    .groups = "drop"
  )

ggplot(plot_df, aes(x = week, y = mean_qids, colour = treatment)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  geom_ribbon(
    aes(ymin = mean_qids - 1.96 * se, ymax = mean_qids + 1.96 * se, fill = treatment),
    alpha = 0.15,
    colour = NA
  ) +
  scale_x_continuous(breaks = c(2, 4, 8, 12, 24, 36, 48)) +
  labs(
    title = "Fake repeated QIDS-SR16 trajectories",
    x = "Week",
    y = "Mean QIDS-SR16",
    colour = "Treatment",
    fill = "Treatment"
  ) +
  theme_minimal(base_size = 12)
```

## 10. Short interpretation

This is what the simulated output is intended to show:

- the two groups are reasonably balanced at baseline
- the pramipexole arm shows a larger improvement by week 12
- the difference narrows by week 48
- the mixed model recovers an adjusted treatment contrast at week 12 that is plausibly in the direction of the published paper

Again, this is only a **didactic fake-data exercise** built to resemble the article, not a reconstruction of the real data.
